---
title: "SurvivalAnalysisCopenhagenStrokeStudy"
author: "Duc-Anh Nguyen"
date: "2025-02-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## EDA
The *cost* dataset from package *pec* collects information from the Copenhagen Stroke Study from 1991 to 1993 (3 years)
```{r}
#install.packages("pec")
library(pec)
library(tidyverse)

data(cost)
head(cost)
skimr::skim(cost)
```

*status* is the variable to predict
*time* measures the length each patient survived in days, value range from 1 to 4262 (11.5 years) 
```{r}
cost <- cost |> select(time, status, age, sex, alcohol, diabetes, smoke)
dim(cost)
prop.table(table(cost$status))

colnames(cost)
skimr::skim(cost)
```
There are 518 patients in this dataset, 22% from them survived after this study, so their data are right censored.
Time and age are the only 2 numerical covariates, the rest are binary

Let us visualise what we have: we try to see what would impact the survival time (in y axis)
```{r}
# install.packages("gridExtra")
# age

plot1 <-
  ggplot(mapping = aes(
    x = age,
    y = time,
    color = as.factor(status)
  ),
  data = cost) + geom_point() + theme() + ggtitle("Age") + xlab("Age") + ylab("Time in days") + scale_color_discrete(name = "Event", labels = c("0", "1"))

# sex
plot2 <- ggplot(mapping = aes(x = sex, y = time, fill = sex), data = cost) +
  geom_boxplot() + theme() + ggtitle("Sex") + xlab("Sex") +
  ylab("Time in days") + theme(legend.position = "none")

# alcohol
plot3 <- ggplot(mapping = aes(x = alcohol, y = time, fill = alcohol),
                data = cost) +
  geom_boxplot() + theme() + ggtitle("Alcohol consumption") + xlab("Alcohol consumption") +
  ylab("Time in days") + theme(legend.position = "none")

# diabetes
plot4 <- ggplot(mapping = aes(x = diabetes, y = time, fill = diabetes),
                data = cost) +
  geom_boxplot() + theme() + ggtitle("Diabetes") +
  xlab("Diabetes") + ylab("Time in days") +
  theme(legend.position = "none")

# smoke
plot5 <- ggplot(mapping = aes(x = smoke, y = time, fill = smoke),
                data = cost) +
  geom_boxplot() + theme() + ggtitle("Smoker") + xlab("Smoker") +
  ylab("Time in days") + theme(legend.position = "none")

gridExtra::grid.arrange(plot1, plot2, plot3, plot4, plot5, nrow = 2)

```
+ Age: there is not really much of a pattern here. All the patients younger than 40 yo survived the study
+ Sex: not much different between the two sexes
+ With diabetes the survival time is shorter, meanwhile consuming alcohol showing a longer survival time. For smoker the median survival time is no difference from non-smoker.


Now visualising with Kaplan-Meier   
```{r}
library(survival)
library(survminer)
km <- survfit(formula = Surv(time = time, event = status) ~ 1, type = "kaplan-meier", data = cost)
km
ggsurvplot(fit = km, data = cost, conf.int = TRUE, surv.median.line = "hv", xlab = "t", ylab = "S(t)", legend = "none", title = "Kaplan-Meier-Estimation")
```
50% the patients survived till day 1794 (bit less than 5 years)

Now let us visualise more focusly on each covariates: age, sex, diabtetes, alcohol and smoke. For age we will bin 2 groups: younger than median and older
```{r}
plots <- list()

# age
cost <-
  cost |> mutate(age_med = ifelse(
    test = age > median(age),
    yes = "yes",
    no = "no"
  ))
km_age_med <-
  survfit(
    formula = Surv(time = time, event = status) ~ age_med,
    type = "kaplan-meier",
    data = cost
  )
km_age_med
plots[[1]] <-
  ggsurvplot(
    fit = km_age_med,
    data = cost,
    surv.median.line = "hv",
    xlab = "t",
    ylab = "S(t)",
    legend = "bottom",
    legend.title = "",
    title = "Age groups"
  )




# sex:
km_sex <- survfit(
  formula = Surv(time = time, event = status) ~ sex,
  type = "kaplan-meier",
  data = cost
)
km_sex
plots[[2]] <- ggsurvplot(
  fit = km_sex,
  data = cost,
  surv.median.line = "hv",
  xlab = "t",
  ylab = "S(t)",
  legend = "bottom",
  legend.title = "",
  title = "Geschlecht"
)




# alcohol consumption:
km_alcohol <- survfit(
  formula = Surv(time = time, event = status) ~ alcohol,
  type = "kaplan-meier",
  data = cost
)
km_alcohol
plots[[3]] <-
  ggsurvplot(
    fit = km_alcohol,
    data = cost,
    surv.median.line = "hv",
    xlab = "t",
    ylab = "S(t)",
    legend = "bottom",
    legend.title = "",
    title = "Alcohol consumption"
  )


# diabetes:
km_diabetes <- survfit(
  formula = Surv(time = time, event = status) ~ diabetes,
  type = "kaplan-meier",
  data = cost
)
km_diabetes
plots[[4]] <- ggsurvplot(
  fit = km_diabetes,
  data = cost,
  surv.median.line = "hv",
  xlab = "t",
  ylab = "S(t)",
  legend = "bottom",
  legend.title = "",
  title = "Diabetes"
)


# Smoker:
km_smoke <- survfit(
  formula = Surv(time = time, event = status) ~ smoke,
  type = "kaplan-meier",
  data = cost
)
km_smoke
plots[[5]] <-
  ggsurvplot(
    fit = km_smoke,
    data = cost,
    surv.median.line = "hv",
    xlab = "t",
    ylab = "S(t)",
    legend = "bottom",
    legend.title = "",
    title = "Smoker"
  )

arrange_ggsurvplots(plots, nrow = 2, ncol = 3)
```
The Kaplan-Meier Curves allign with our initial visualisations
+ Age: the younger group has median survival time 2 as long as the older (2233 vs 1169 days)
+ Just as our first visualisation, the survival chance of people without *diabetes* is higher than people with, and higher for people who drinks alcohol than not drinking
+ *Smoker*'s chance of survival is not much of a diffence from non-smoker, but only till day 1800, after that smoker has a slightly higher chance of suvive

With Kaplan-Meier curve we can only see effect of each covariate. Now we want to fit a regression model for multivariate analysis!



## Analysis of suvival time of stroke patient with *Cox model*
```{r}
model <- survival::coxph(formula = Surv(time, status)~age + sex + alcohol + diabetes + smoke, data = cost)
summary(model)
```
+ Age: the hazard rate λ(t) (the risk in the consecutive moment t+1 to die when one survive till t) rises avg. exponentitallly exp(0.055) = 1.06 for each extra year of surviving, c.p.

+ Sex: for men the rate rises every year extra 1.4014 compared to women, c.p.
+ Alcohol: people using alcohol have a lower hazard rate than not using (0.9174 times as high), c.p.

+ Diabetes: With diabetes is the rate 1.5542 times higher than without, c.p.

+ Smoker: Non-Smoker have a higher hazard rate (1.2931 times higher than smoker) c.p.

